---
title: "Final Project Air Travel"
author: "Juliet Cramer & Elie Rizk"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
##knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(transformr)    # for "tweening" (gganimate)
library(readr)
library(scales)
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```


```{r}
number_of_commercial_flights <- read_csv("number_of_commercial_flights.csv", col_types = cols(DateTime = col_character())) #This is the flight data set

Fixed_Unemployment <- read_csv("Fixed Unemployment.csv", 
    skip = 1)

TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))
#This is the data set for TSA Travelers 

#COVID-19 data from the New York Times
covid19 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

#Commercial Flights data
number_of_commercial_flights <- read_csv("number_of_commercial_flights.csv", col_types = cols(DateTime = col_character()))

#TSA Travelers data
TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))

UNRATE_Large <- read_csv("UNRATE_Large.csv")

Number_Of_Flights <- read_csv("Number_Of_Flights.csv")
```

## R Markdown


```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2020 through TSA", 
       x = " ", 
       y = " ")
```

```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav_Year_Ago`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2019 through TSA", 
       x = " ", 
       y = " ")
```

## Including Plots

You can also embed plots, for example:

I want to explore a little bit as to how the unemployment rate and the traffic through TSA are related. Is there a correlation there? My hypothesis is that unemployment rates and TSA traffic are inversely related. In other words, when unemployment is high (such as during the COVID-19 pandemic), the number of people traveling is low. Let's see. 

```{r}
Number_Of_Flights %>% 
  group_by(Year) %>% 
  summarize(total_travel = sum(Total)) %>%
  ggplot(aes(y = total_travel, 
             x = Year)) + 
  geom_point() + 
  scale_y_continuous(labels = comma) + 
  labs(title = "Total Number of Flights from 2007-2008 and 2018-2020")
```

```{r}

```

