---
title: "Final Project Air Travel"
author: "Juliet Cramer"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(readr)
```


```{r}
number_of_commercial_flights <- read_csv("number_of_commercial_flights.csv", col_types = cols(DateTime = col_character())) #This is the flight data set
Unemployment_Rate <- read_csv("Unemployment_Rate.csv", skip = 2)
#This is the unemployment rate Data Set
TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))
#This is the data set for TSA Travelers
```

## R Markdown


```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2020 through TSA", 
       x = " ", 
       y = " ")
```

```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav_Year_Ago`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2019 through TSA", 
       x = " ", 
       y = " ")
```

## Including Plots

You can also embed plots, for example:

I want to explore a little bit as to how the unemployment rate and the traffic through TSA are related. Is there a correlation there? My hypothesis is that unemployment rates and TSA traffic are inversely related. In other words, when unemployment is high (such as during the COVID-19 pandemic), the number of people traveling is low. Let's see. 

```{r}
Unemployment_Rate %>% ##NEEDS WORK
  pivot_longer(cols = Jan:Dec, 
              names_to = "Month")
```

