---
title: "Why has flight travel been impacted?"
author: "Juliet Cramer, Elie Rizk and Julius Enarsson Enestrom"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
##knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(transformr)    # for "tweening" (gganimate)
library(readr)
library(scales)      # for even more plotting themes
library(ggplot2)
library(ggtext)    
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```


```{r}
#number_of_commercial_flights <- read_csv("number_of_commercial_flights.csv", col_types = cols(DateTime = col_character())) #This is the flight data set

#Fixed_Unemployment <- read_csv("Fixed Unemployment.csv", DONT THINK WE NEED THIS FILE, CAN USE UNRATE_Large instead
  #  skip = 1)

TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))
#This is the data set for TSA Travelers 

#COVID-19 data from the New York Times
covid19 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

#Commercial Flights data
#flight_data_2020 <- read_csv("number_of_commercial_flights.csv", col_types = cols(DateTime = col_character()))

#TSA Travelers data
TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))

unrate_large <- read_csv("UNRATE_Large.csv")

number_of_passengers <- read_csv("Number_Of_Flights.csv")
```

## R Markdown


```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2020 through TSA", 
       x = " ", 
       y = " ")
```

```{r}
TSA_Travelers_Data_2020 %>% ##NEEDS WORK HELP!
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot(aes(x = Date,
             y = `Tot_Trav_Year_Ago`)) +
  geom_line() +
  labs(title = "Total Number of Travelers by Month 2019 through TSA", 
       x = " ", 
       y = " ")
```

## Including Plots

You can also embed plots, for example:

I want to explore a little bit as to how the unemployment rate and the traffic through TSA are related. Is there a correlation there? My hypothesis is that unemployment rates and TSA traffic are inversely related. In other words, when unemployment is high (such as during the COVID-19 pandemic), the number of people traveling is low. Let's see. 



```{r}
TSA_Travelers_Data_2020 %>% 
  arrange(Date) %>% 
  rename(Tot_Trav_Year_Ago = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         Tot_Trav = `Total Traveler Throughput`) %>% 
  ggplot() +
  geom_line(aes(x= Date, y = Tot_Trav), color = "blue") +
  geom_line(aes(x = Date, y = Tot_Trav_Year_Ago), color = "red") +
  geom_smooth(aes(x = Date, 
                  y = Tot_Trav_Year_Ago),
                  color = "firebrick4",
                  se = FALSE)+
  geom_smooth(aes(x = Date, 
                  y = Tot_Trav),
                  color = "dodgerblue4",
                  se = FALSE)+
  scale_y_continuous(labels = scales::comma) +
  labs(title = "TSA Travelers 2019 vs. 2020", 
       x = " ", 
       y = " ")  +
 #      subtitle = "Date: {frame_along}") +
#  transition_reveal(Date)+
#  exit_fade()+
  theme(legend.title = element_blank(),
        plot.background = element_rect(fill = "#b6d4c8"),
        axis.title = element_blank(),
        plot.title = element_text(size = "15", face = "bold")) #-> travelers

#travelers_ggplotly <- ggplotly(
#  p = travelers,
 # width = NULL,
  #height = NULL,
  #tooltip = "all",
  #dynamicTicks = FALSE,
  #layerData = 1,
  #originalData = TRUE,
  #source = "A"
#)

#htmlwidgets::saveWidget(travelers_ggplotly, "travelers_ggplotly.html")

#travelers_ggplotly
```

```{r}
covid19 %>% 
  group_by(date) %>% 
  summarise(total_num_cases = sum(cases), total_num_death = sum(deaths)) %>% 
  ggplot() +
  geom_line(aes(x = date, y = total_num_cases, color = "blue")) +
  #geom_line(aes(x = date, y = total_num_death, color = "red")) + 
  labs(x = "",
       y = "",
       color = "",
       title = "Cumulative COVID-19 Cases and Deaths in US") + 
  theme(legend.position = "bottom") +
  scale_y_log10(labels = comma) + #uses the scales library
  scale_x_date(date_breaks = "1 month",
               date_labels = "%b",
               date_minor_breaks = "1 week")
```

```{r}
Year <- as.numeric(as.character(number_of_passengers$Year))

number_of_passengers %>%
  group_by(Year, Month) %>% 
  summarize(Month, Total) %>% 
  filter(Year %in% c("2007", "2008")) %>% 
  ggplot(aes(x = Month, 
             y = Total,
             color = Year,
            group_vars())) +
  geom_point() + 
  geom_path(aes(group = Year)) +
  scale_color_continuous_tableau() +
  scale_y_continuous(labels = comma) 
```

```{r}
number_of_passengers %>%
  filter(Year == 2008) %>% 
  ggplot(aes(x = Month, 
             y = Total)) + 
  geom_line() + 
  scale_y_continuous(labels = comma) 
```
```{r}
number_of_passengers %>%
  filter(Year == 2019) %>% 
  ggplot(aes(x = Month, 
             y = Total)) + 
  geom_line() + 
  scale_y_continuous(labels = comma) 
```

```{r}
number_of_passengers %>%
  filter(Year == 2020) %>% 
  ggplot(aes(x = Month, 
             y = Total)) + 
  geom_line() + 
  scale_y_continuous(labels = comma) 
```

```{r}
unrate_large %>% 
  filter(between(DATE, as.Date("2019-01-01"), as.Date("2020-11-01"))) %>%
  ggplot() + 
  geom_line(aes(x = DATE, y = UNRATE))
```


```{r}
number_of_passengers %>%
  group_by(Year) %>% 
  summarize(total_travel = sum(Total)) %>%
  ggplot(aes(y = total_travel, 
             x = Year)) + 
  geom_point() + 
  scale_y_continuous(labels = comma) + 
  labs(title = "Total Number of Flights from 2007-2008 and 2018-2020")
```
