---
title: "Why has flight travel been impacted?"
author: "Juliet Cramer, Elie Rizk and Julius Enarsson Enestrom"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
##knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(base)
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(transformr)    # for "tweening" (gganimate)
library(readr)
library(scales)      # for even more plotting themes
library(ggplot2)
library(ggtext)    
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```


```{r}
#This is the data set for TSA Travelers 
TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))

#COVID-19 data from the New York Times
covid19 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

#TSA Travelers data
TSA_Travelers_Data_2020 <- read_csv("TSA Travelers Data 2020.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))

unrate_large <- read_csv("UNRATE_Large.csv") %>% 
  mutate(MonthAbb = month(DATE,label = TRUE))

number_of_passengers <- read_csv("Number_Of_Flights.csv")

number_of_passengers <- transform(number_of_passengers, MonthAbb = month.abb[Month])
```

## R Markdown


## Including Plots

You can also embed plots, for example:

I want to explore a little bit as to how the unemployment rate and the traffic through TSA are related. Is there a correlation there? My hypothesis is that unemployment rates and TSA traffic are inversely related. In other words, when unemployment is high (such as during the COVID-19 pandemic), the number of people traveling is low. Let's see. 


Graph #1 

```{r}
TSA_Travelers_Data_2020 %>% 
  arrange(Date) %>% 
  rename(`Total Travelers 2019` = `Total Traveler Throughput (1 Year Ago - Same Weekday)`, 
         `Total Travers 2020` = `Total Traveler Throughput`) %>% 
  ggplot() +
  geom_line(aes(x= Date, y = `Total Travers 2020`), color = "dodgerblue4", alpha = 0.7) +
  geom_line(aes(x = Date, y = `Total Travelers 2019`), color = "firebrick4", alpha = 0.7) +
  geom_smooth(aes(x = Date, 
                  y = `Total Travelers 2019`),
                  color = "firebrick4",
                  se = FALSE)+
  geom_smooth(aes(x = Date, 
                  y = `Total Travers 2020`),
                  color = "dodgerblue4",
                  se = FALSE)+
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")+
  labs(title = "Number of People Traveling by Plane in <b style='color:#970111'>2019</b> vs. <span style='color:#004e8f;'>2020</span>", 
       x = " ", 
       y = " ",
       subtitle = "Data for US",
       caption = "Data from TSA") +
  theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "gray95", color = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        plot.title = element_markdown(size = "15", face = "bold"),
        panel.grid.major.y = element_line(color = "gray70"),
        panel.grid.major.x = element_line(color = "gray70"),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
```

Graph #2

```{r}
covid19 %>% 
  group_by(date) %>% 
  summarise(total_num_cases = sum(cases), total_num_death = sum(deaths)) %>% 
  ggplot() +
  geom_line(aes(x = date, 
                y = total_num_cases), color = "dodgerblue4", alpha = 0.7) +
  geom_line(aes(x = date, 
                y = total_num_death), color = "firebrick4", alpha = 0.7) + 
  labs(x = "",
       y = "",
       color = "",
       title = "2020 Cumulative COVID-19 <b style='color:#970111'>Deaths</b> and <span style='color:#004e8f;'>Cases</span> in U.S.", 
       caption = "Data from NYT") + 
  scale_y_log10(labels = comma) +
  scale_x_date(date_breaks = "1 month",
               date_labels = "%b",
               date_minor_breaks = "1 week")+
   theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "gray95", color = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        plot.title = element_markdown(size = "15", face = "bold"),
        panel.grid.major.y = element_line(color = "gray70"),
        panel.grid.major.x = element_line(color = "gray70"),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
```


Graph #3

```{r}
number_of_passengers %>%
  filter(Year %in% c("2007", "2008")) %>% 
  mutate(year = ifelse(Year < 2008, 
                           "2007", "2008")) %>%
  rename(`Total Passengers` = Total) %>%
  ggplot(aes(x = fct_inorder(MonthAbb), 
             y = `Total Passengers`,
             color = year)) +
  geom_point() + 
  geom_path(aes(group = Year)) +
  scale_color_manual(values = c("#970111", "#004e8f")) +
  scale_y_continuous(labels = scales::comma)+
  labs(title = "Number of Passengers <b style='color:#970111'>2007</b> vs. <span style='color:#004e8f;'>2008</span>", 
       x = " ", 
       y = " ", 
       caption = "Data from U.S. BTS") +
  theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "gray95", color = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        plot.title = element_markdown(size = "15", face = "bold"),
        panel.grid.major.y = element_line(color = "gray70"),
        panel.grid.major.x = element_line(color = "gray70"),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        legend.position = "none")
```

Graph #4

```{r}
number_of_passengers %>%
  filter(Year %in% c("2019", "2020")) %>% 
  mutate(year = ifelse(Year < 2020, 
                           "2019", "2020")) %>% 
  rename(`Total Passengers` = Total) %>%
  ggplot(aes(x = fct_inorder(MonthAbb),
             y = `Total Passengers`,
             color = year)) +
  geom_point() + 
  geom_path(aes(group = Year)) +
  scale_color_manual(values = c("#970111", "#004e8f")) +
  scale_y_continuous(labels = comma)+
  labs(title = "Number of Passengers <b style='color:#970111'>2019</b> vs. <span style='color:#004e8f;'>2020</span>", 
       x = " ", 
       y = " ", 
       caption = "Data from U.S. BTS") +
   theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "gray95", color = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        plot.title = element_markdown(size = "15", face = "bold"),
        panel.grid.major.y = element_line(color = "gray70"),
        panel.grid.major.x = element_line(color = "gray70"),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        legend.position = "none")
```


Graph #5

```{r}
unrate_large %>% 
  filter(between(DATE, as.Date("2020-01-01"), as.Date("2020-11-01")) | between(DATE, as.Date("2008-01-01"), as.Date("2008-12-31"))) %>%
  mutate(Year = ifelse(DATE < "2010-01-01", 
                           "2008", "2020")) %>% 
  select(DATE, UNRATE, MonthAbb, Year) %>%
  rename(Month = MonthAbb) %>% 
  ggplot(aes(x = Month, y = UNRATE, color = Year, group = Year)) + 
  geom_point() +
  geom_line() +
  scale_color_manual(values = c("#970111", "#004e8f")) +
  labs(title = "Unemployment Rate (%) <b style='color:#970111'>2008</b> vs. <span style='color:#004e8f;'>2020</span>", 
       x = " ", 
       y = " ", 
       caption = "Data from Federal Reserve Bank (FRED)") +
  theme(legend.title = element_blank(),
        panel.background = element_rect(fill = "gray95", color = "gray95"),
        plot.background = element_rect(fill = "gray95"),
        plot.title = element_markdown(size = "15", face = "bold"),
        panel.grid.major.y = element_line(color = "gray70"),
        panel.grid.major.x = element_line(color = "gray70"),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        legend.position = "none")
```




